Exact expressions for double descent and implicit regularization \\ via surrogate random design

### 2022-08-23 design/continuous_colt/continuous.tex
Commit original repos

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathabx}
\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{xfrac}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}
\usepackage{cleveref}
\input{shortdefs}

@@

% \author{
%       Micha{\l } Derezi\'{n}ski \\
% Department of Statistics\\
% University of California, Berkeley\\
% \texttt{mderezin@berkeley.edu}
% }

% \setlength{\parindent}{0cm}
% \setlength{\parskip}{0mm}

@@

@@

Classical statistical learning theory asserts that to achieve generalization one must use training sample size that sufficiently exceeds the complexity of the learning model, where the latter is typically represented by the number of parameters (or some related structural parameter) \cite{HFT09}. 
In particular, this seems to suggest the conventional wisdom that one should not use models that fit the training data exactly. 
However, modern machine learning
practice often seems to go against this intuition, using models with so many
parameters that the training data can be perfectly interpolated, in which case
the training error vanishes. It has been shown that 
models such as deep neural networks, as well as certain so-called interpolating kernels and decision
trees, can generalize well in this regime. In particular, recent work \cite{BHMM19}
empirically demonstrated a phase transition in generalization
performance of learning models which occurs at an
\emph{interpolation thershold}, i.e., a point where training error
goes to zero (as one varies the ratio between the model complexity and
the sample size). Moving away from this threshold in 
either direction tends to reduce the generalization error, leading to
the so-called \emph{double descent} curve.

To understand this
surprising phenomenon, in perhaps the simplest possible setting, we
study it in the context of linear or least squares regression.
Consider a full rank $n\times d$ data matrix $\X$ and a vector $\y$ of
responses corresponding to each of the $n$ data points (the rows of $\X$), where we wish to
find the best linear model $\X\w\approx \y$, parameterized by a
$d$-dimensional vector $\w$.
The simplest example of an estimator that has been shown to exhibit
the double descent phenomenon \cite{belkin2019two} is the
Moore-Penrose estimator, $\wbh=\X^\dagger\y$: 
in the so-called over-determined regime, i.e., when $n>d$, it corresponds to the
least squares solution, i.e., $\argmin_{\w} \|\X\w-\y\|^2$; and in the
under-determined regime (also known as
over-parameterized or interpolating), i.e., when $n<d$, it
corresponds to the minimum norm solution to the linear system $\X\w=\y$.  
Given the ubiquity of linear regression and the Moore-Penrose
solution, e.g., in kernel-based machine learning, studying the
performance of this estimator can shed some light on the effects of
over-parameterization/interpolation in machine learning more generally. 
Of particular interest are results that are exact (i.e., not upper/lower bounds) and
non-asymptotic (i.e., for large but still finite $n$ and $d$).

We build on methods from Randomized Numerical
Linear Algebra (RandNLA) in order to obtain \emph{exact non-asymptotic
expressions} for the mean squared error (MSE) of the
Moore-Penrose estimator (see Theorem~\ref{t:mse}). 
This provides a precise characterization of the double descent
phenomenon for perhaps the simplest and most ubiquitous regression
problem. 
In obtaining these results, we are able to provide precise
formulas for the \emph{implicit regularization} induced by
minimum norm solutions of
under-determined training samples, relating it to classical ridge regularization (see Theorem~\ref{t:unbiased}). 
This result has been observed empirically for RandNLA
methods~\cite{Mah-mat-rev_JRNL}, but it has also been shown 
in deep learning~\cite{Ney17_TR} and machine
learning~\cite{Mah12} more generally. 
To obtain our precise results, we use a somewhat non-standard random
design, which we term surrogate random design (see
Section~\ref{s:determinantal} for a detailed discussion), and which we
expect will be of more general interest. 
Informally, the goal of a surrogate random design is to modify an
original design to capture its main properties
while being ``nicer'' in some useful way. In
Theorem~\ref{t:asymptotic} and Section \ref{s:asymptotic} we show,
both theoretically and empirically, that our surrogate design
accurately preserves the key properties of the original design when
the data distribution is a multivariate Gaussian.

@@

As the performance metric in our analysis, we use the \emph{mean
  squared error} (MSE), defined as
$\mathrm{MSE}[\wbh]=\E\big[\|\wbh-\w^*\|^2\big]$, where $\w^*$ is a fixed
underlying linear model of the responses.  
In analyzing the MSE, we make the following standard assumption on the response noise.
\begin{assumption}[Homoscedastic noise]\label{a:linear}
  Responses are $y(\x) = \x^\top\w^*+\xi$ where
$\xi\sim\Nc(0,\sigma^2)$.
\end{assumption}

\noindent
Our main result provides an exact expression for the MSE of the
Moore-Penrose estimator under our surrogate design denoted $S_\mu^n$, where
$\mu$ is the $d$-variate distribution of the row vector $\x^\top$ and $n$ is the sample
size (details in Section~\ref{s:determinantal}). This surrogate is
used in place of the standard $n\times d$ random design $\X\sim\mu^n$, where $n$
data points (the rows of $\X$) are sampled independently from
$\mu$. Unlike for the standard design, our MSE formula is
fully expressible as a function of the covariance
matrix $\Sigmab_\mu=\E_\mu[\x\x^\top]$. To state our main result, we
need an additional minor assumption on $\mu$ which is satisfied by
most standard continuous distributions such as any multivariate
Gaussian with positive definite covariance matrix. 
\begin{assumption}[General position]\label{a:general-position}
For $1\leq n \leq d$, if $\X\sim\mu^n$, then $\rank(\X)=n$ almost surely.
\end{assumption}

\noindent
Under Assumptions~\ref{a:linear} and~\ref{a:general-position}, we can establish our first main result, stated as the following theorem.

\begin{theorem}[Exact non-asymptotic MSE]
\label{t:mse}
  If the response noise is homoscedastic with variance~$\sigma^2$
  (Assumption~\ref{a:linear}) and $\mu$ is in
  general position (Assumption~\ref{a:general-position}), then for
  $\Xb\sim S_\mu^n$ (Definition~\ref{d:main}) and
  $\yb_i=y(\xbb_i)$, we have
  \begin{align*}
    \MSE{\Xb^\dagger\ybb} =
    \begin{cases}
    \sigma^2\,\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)\cdot
    \frac{1-\alpha_n}{d-n}\ +\
\frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
{\tr((\Sigmab_\mu+\lambda_n\I)^{-1})}\cdot (d-n),
&\text{for }n<d,\\
\sigma^2\, \tr(\Sigmab_\mu^{-1}),& \text{for }n=d,\\
\sigma^2\,\tr(\Sigmab_\mu^{-1})\cdot\frac{1-\beta_n}{n-d},&\text{for
}n>d,
\end{cases}
  \end{align*}
  where $\lambda_n\geq 0$ satisfies
  $\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})=n$, while
  $\alpha_n=\det(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$
  and $\beta_n=\ee^{d-n}$.
\end{theorem}
\begin{definition}
  We will use $\Mc(\Sigmab_\mu, \w^*,\sigma^2,n)$ to denote the above expressions
  for $\MSE{\Xb^\dagger\ybb}$.
\end{definition}
For illustration, we plot these MSE expressions in Figure \ref{f:intro}a,
comparing them with empirical estimates of the true MSE under the
i.i.d.~design for a multivariate Gaussian distribution
$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
$100$ and vary the number of samples $n$, observing a double descent
peak at $n=d$. We observe that our theory aligns well with
the empirical estimates, whereas
previously, no such theory was available except for special
cases such as $\Sigmab=\I$ (more details in Theorem \ref{t:asymptotic}
and Section \ref{s:asymptotic}). The plots
show that varying the spectral decay of $\Sigmab$ has a significant effect on the
shape of the curve in the under-determined regime. We use the
horizontal line to denote the MSE of the null estimator
$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
decay rapidly, then the Moore-Penrose estimator suffers less error 
than the null estimator for some values of $n<d$, and the curve
exhibits a local optimum in this regime.

\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.48\textwidth}
%  \begin{center}
    \includegraphics[width=\textwidth]{figs/descent-intro}
%  \end{center}
  \caption{
Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
numerical estimates even for non-isotropic
features. Eigenvalue decay leads to a steeper 
descent curve in the under-determined regime ($n<d$).}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.48\textwidth}
%  \begin{center}
    \includegraphics[width=\textwidth]{figs/descent-shrinkage}
%  \end{center}
  \caption{
    The mean of the estimator $\X^\dagger\y$ exhibits
    shrinkage which closely matches the shrinkage of a
    ridge-regularized least squares optimum (theory lines), as characterized by
    Theorem \ref{t:unbiased}.}
\end{subfigure}
\caption{Illustration of the main results for $d=100$ and
$\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
eigenvalues decaying exponentially and scaled so that
$\tr(\Sigmab^{-1})=d$. We use our surrogate
formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
\ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
lines), accompanied by the empirical estimates based on the standard
i.i.d.~design (error bars are three times the standard error of the
mean). We consider three different condition numbers $\kappa$ of
$\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
i.e., $\Sigmab=\I$. In all cases, we use $\sigma^2=\|\w^*\|=1$ with
$\w^*=\frac1{\sqrt{d}}\one$.}
\label{f:intro}
\end{figure}

One important aspect of Theorem~\ref{t:mse} comes from the relationship between $n$ and the parameter $\lambda_n$, which together satisfy $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$. 
This expression is precisely the classical notion of \emph{effective dimension} for ridge regression regularized with $\lambda_n$~\cite{Alaoui15}, and it arises here even though there is no explicit ridge regularization in the problem being considered in Theorem~\ref{t:mse}. 
The global solution to the ridge regression task (i.e., $\ell_2$-regularized
least squares) with parameter $\lambda$ is defined as: 
\begin{align*}
\argmin_\w \Big\{\E_{\mu,y}\big[\big(\x^\top\w-y(\x)\big)^2\big]
    + \lambda\|\w\|^2\Big\}\ =\ (\Sigmab_\mu +
  \lambda\I)^{-1}\v_{\mu,y},\qquad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
\end{align*}
When Assumption \ref{a:linear} holds, then
$\v_{\mu,y}=\Sigmab_\mu\w^*$, however ridge-regularized least squares
is well-defined for much more general response models.
Our second result makes a direct connection between the (expectation
of the) unregularized minimum norm solution on the sample
and the global ridge-regularized solution.
While the under-determined regime (i.e., $n<d$) is of primary interest to us, 
for completeness we state this result for arbitrary values of $n$ and $d$.
Note that, just like the definition of regularized least squares, this
theorem applies more generally than Theorem~\ref{t:mse}, in that it
does \emph{not} require the responses to follow any linear model as in Assumption~\ref{a:linear}.  
\begin{theorem}[Implicit regularization of Moore-Penrose estimator]
\label{t:unbiased}
For $\mu$ satisfying\footnote{The proof of Theorem \ref{t:unbiased}
  can be easily extended to probability measures $\mu$ that do not
  satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
  assumption here to simplify the presentation.} Assumption~\ref{a:general-position} and any
  $y(\cdot)$ such that   $\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined, if
  $\Xb\sim S_\mu^n$ and $\yb_i=y(\xbb_i)$, then
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big] = 
    \begin{cases}
       (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} &\text{for }n<d,\\
        \Sigmab_\mu^{-1}\v_{\mu,y} &\text{for }n \ge d,
    \end{cases}
  \end{align*}
  where, as in Theorem \ref{t:mse}, $\lambda_n$ is chosen so that the effective dimension
  $\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$ equals $n$.
\end{theorem}

@@

We illustrate this result in Figure
\ref{f:intro}b, plotting the norm of the expectation of the
Moore-Penrose estimator. As for the MSE, our surrogate theory aligns
well with the empirical estimates for i.i.d.~Gaussian designs, showing
that the shrinkage of the unregularized estimator in the
under-determined regime matches the implicit
ridge-regularization characterized by Theorem \ref{t:unbiased}. While the shrinkage
is a linear function of the sample size $n$ for isotropic features
(i.e., $\Sigmab=\I$), it 
exhibits a non-linear behavior for other spectral decays.
Such \emph{implicit regularization} has been studied
previously~\cite{MO11-implementing, PM11, GM14_ICML, Mah12}; it has
been observed empirically for RandNLA sampling
algorithms~\cite{Mah-mat-rev_JRNL, MMY15}; and it has also received attention more
generally within the context of neural networks~\cite{Ney17_TR}. While our implicit regularization result
is limited to the Moore-Penrose estimator, this new connection (and
others, described below) between the minimum norm solution of an unregularized
under-determined system and a ridge-regularized least squares solution
offers a simple interpretation for the implicit regularization
observed in modern machine learning architectures.

Our exact non-asymptotic expressions in Theorem~\ref{t:mse} and 
our exact implicit regularization results in Theorem~\ref{t:unbiased}
are derived for the surrogate design, but Figure \ref{f:intro}
suggests that they accurately describe the MSE (up to lower order
terms) also under the standard i.i.d.~design $\X\sim\mu^n$,
particularly when $\mu$ is a multivariate Gaussian.  
As a third result, we can verify this in the cases where there exist
known expressions for the MSE under the i.i.d.~design (standard
Gaussian for the under-determined setting, and 
arbitrary Gaussian for the over-determined one).
\begin{theorem}[Asymptotic consistency of surrogate design]
\label{t:asymptotic}
Let $\rho=n/d\neq 1$, $\X\sim \mu^n$
and $y_i=y(\x_i)$ satisfy Assumption~\ref{a:linear}. If $d\geq
c_\rho=\frac3{|1-\rho|}$ and
\begin{align*}
  \mu &= \begin{cases}
    \Nc(\zero,\I)&\text{when }n<d-1,\\
    \Nc(\zero,\Sigmab),\ \Sigmab\succ\zero &\text{when }n>d+1,
  \end{cases}
\end{align*}
then the absolute difference between surrogate
expressions and the true MSE is bounded as follows:
% &\text{if}\qquad  (1)\quad \rho<1\ \text{ and }\ \mu=\Nc(\zero,\I)\qquad\text{or}\qquad
%     (2)\quad \rho > 1\ \text{ and }\ \mu=\Nc(\zero,\Sigmab),\
%     \Sigmab\succ\zero,\\
\begin{align*}
\Big|\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)\Big|\ \leq\
\frac{c_\rho}{d} \cdot \Mc(\Sigmab,\w^*,\sigma^2,n).
\end{align*}
\end{theorem}
\begin{remark}
For $n$ equal to $d-1$, $d$ or $d+1$, the true MSE under Gaussian
random design can be infinite, whereas the
surrogate MSE is finite and has a closed form expression.
\end{remark}

\noindent
Empirical estimates given in Figure \ref{f:intro} suggest that the
consistency of surrogate expressions holds much more generally than
it is stated above. Based on a detailed empirical analysis described in
Section~\ref{s:empirical}, we conjecture that an asymptotic consistency
result of the form similar to the statement of Theorem~\ref{t:asymptotic}
holds true in the under-determined regime without the assumption that
$\Sigmab=\I$. In this case, no formula is known for
$\MSE{\X^\dagger\y}$, whereas the expressions for the surrogate
Gaussian design naturally extend.

>>>>

\newif\ifisarxiv
\isarxivtrue

\ifisarxiv
\documentclass[11pt]{article}
\usepackage[round]{natbib}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathabx}
\usepackage{color}
\usepackage{cancel}
\usepackage{wrapfig}
\usepackage{subfigure}
\usepackage{caption}
\usepackage{xfrac}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}
\usepackage{cleveref}

@@

%  \newcommand{\jmlrBlackBox}{\rule{1.5ex}{1.5ex}}
  \providecommand{\BlackBox}{\rule{1.5ex}{1.5ex}}
%  \newcommand{\jmlrQED}{\hfill\jmlrBlackBox\par\bigskip}
  \renewenvironment{proof}%
  {%
   \par\noindent{\bfseries\upshape Proof\ }%
  }%
  {\hfill\BlackBox\par\bigskip}

\else
\documentclass[anon]{colt2020}

\title[Exact expressions for double descent and implicit
regularization]{Exact expressions for double descent and implicit
  regularization\\
  via surrogate random design}
\usepackage{times}

\coltauthor{%
  \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
  \addr Department of Statistics\\
  University of California, Berkeley\\
  \AND
  \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
  \addr Department of Statistics\\
  University of California, Berkeley\\
  \AND
  \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
  \addr ICSI and Department of Statistics\\
  University of California, Berkeley
}

\usepackage{float}
\usepackage{wrapfig}

\fi

\input{../shortdefs}

@@

\ifisarxiv\else
\begin{keywords}
  Least squares, determinantal point process,
  double descent, implicit regularization
\end{keywords}
\fi

@@

Classical statistical learning theory asserts that to achieve generalization
one must use training sample size that sufficiently exceeds the complexity of
the learning model, where the latter is typically represented by the number of
parameters \citep[or some related structural parameter; see][]{HFT09}.  In particular,
this seems to suggest the conventional wisdom that one should not use models
that fit the training data exactly.  However, modern machine learning practice
often seems to go against this intuition, using models with so many parameters
that the training data can be perfectly interpolated, in which case the
training error vanishes. It has been shown that models such as deep neural
networks, as well as certain so-called interpolating kernels and decision
trees, can generalize well in this regime. In particular,
\cite{BHMM19} empirically demonstrated a phase transition in generalization
performance of learning models which occurs at an \emph{interpolation
thershold}, i.e., a point where training error goes to zero (as one varies the
ratio between the model complexity and the sample size). Moving away from this
threshold in either direction tends to reduce the generalization error, leading
to the so-called \emph{double descent} curve.

To understand this
surprising phenomenon, in perhaps the simplest possible setting, we
study it in the context of linear or least squares regression.
Consider a full rank $n\times d$ data matrix $\X$ and a vector $\y$ of
responses corresponding to each of the $n$ data points (the rows of $\X$), where we wish to
find the best linear model $\X\w\approx \y$, parameterized by a
$d$-dimensional vector $\w$.
The simplest example of an estimator that has been shown to exhibit
the double descent phenomenon \citep{belkin2019two} is the
Moore-Penrose estimator, $\wbh=\X^\dagger\y$:
in the so-called over-determined regime, i.e., when $n>d$, it corresponds to the
least squares solution, i.e., $\argmin_{\w} \|\X\w-\y\|^2$; and in the
under-determined regime (also known as
over-parameterized or interpolating), i.e., when $n<d$, it
corresponds to the minimum norm solution to the linear system $\X\w=\y$.
Given the ubiquity of linear regression and the Moore-Penrose
solution, e.g., in kernel-based machine learning, studying the
performance of this estimator can shed some light on the effects of
over-parameterization/interpolation in machine learning more generally.
Of particular interest are results that are exact (i.e., not upper/lower bounds) and
non-asymptotic (i.e., for large but still finite $n$ and~$d$).

We build on methods from Randomized Numerical Linear Algebra (RandNLA) in order
to obtain \emph{exact non-asymptotic expressions} for the mean squared error
(MSE) of the Moore-Penrose estimator (see Theorem~\ref{t:mse}).  This provides
a precise characterization of the double descent phenomenon for perhaps the
simplest and most ubiquitous regression problem.  In obtaining these results,
we are able to provide precise formulas for the \emph{implicit regularization}
induced by minimum norm solutions of under-determined training samples,
relating it to classical ridge regularization (see Theorem~\ref{t:unbiased}).
This result has been observed empirically for RandNLA
methods~\citep{Mah-mat-rev_JRNL}, but it has also been shown in deep
learning~\citep{Ney17_TR} and machine learning~\citep{Mah12} more generally.  To
obtain our precise results, we use a somewhat non-standard random design, which
we term surrogate random design (see Section~\ref{s:determinantal} for a
detailed discussion), and which we expect will be of more general interest.
Informally, the goal of a surrogate random design is to modify an original
design to capture its main properties while being ``nicer'' in some useful way.
In Theorem~\ref{t:asymptotic} and Appendix \ref{sec:asymp-conj-details} we show, both
theoretically and empirically, that our surrogate design accurately preserves
the key properties of the original design when the data distribution is a
multivariate Gaussian.

@@

As the performance metric in our analysis, we use the \emph{mean
  squared error} (MSE), defined as
$\mathrm{MSE}[\wbh]=\E\big[\|\wbh-\w^*\|^2\big]$, where $\w^*$ is a fixed
underlying linear model of the responses.
In analyzing the MSE, we make the following standard assumption that
the response noise is homoscedastic with variance $\sigma^2$.
\begin{assumption}[Homoscedastic noise]\label{a:linear}
  Responses are $y(\x) = \x^\top\w^*+\xi$ where
$\xi\sim\Nc(0,\sigma^2)$.
\end{assumption}

\noindent
Our main result provides an exact expression for the MSE of the
Moore-Penrose estimator under our surrogate design denoted $\Xb\sim S_\mu^n$, where
$\mu$ is the $d$-variate distribution of the row vector $\x^\top$ and $n$ is the sample
size (details in Section~\ref{s:determinantal}). This surrogate is
used in place of the standard $n\times d$ random design $\X\sim\mu^n$, where $n$
data points (the rows of $\X$) are sampled independently from
$\mu$. Unlike for the standard design, our MSE formula is
fully expressible as a function of the covariance
matrix $\Sigmab_\mu=\E_\mu[\x\x^\top]$. To state our main result, we
need an additional minor assumption on $\mu$ which is satisfied by
most standard continuous distributions such as any multivariate
Gaussian with positive definite covariance matrix.
\begin{assumption}[General position]\label{a:general-position}
For $1\leq n \leq d$, if $\X\sim\mu^n$, then $\rank(\X)=n$ almost surely.
\end{assumption}

\noindent
Under Assumptions~\ref{a:linear} and~\ref{a:general-position}, we can
establish our first main result, stated as the following theorem, where
we use $\X^\dagger$ to denote the Moore-Penrose inverse of $\X$.
% , which gives an exact non-asymptotic
% expression for the MSE of the Moore-Penrose estimator  the
% surrogate design $S_\mu^n$.

\begin{theorem}[Exact non-asymptotic MSE]
\label{t:mse}
  If the response noise is homoscedastic % with variance~$\sigma^2$
  (Assumption~\ref{a:linear}) and $\mu$ is in
  general position (Assumption~\ref{a:general-position}), then for
  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and
  $\yb_i=y(\xbb_i)$, %we have
  \begin{align*}
    \MSE{\Xb^\dagger\ybb} =
    \begin{cases}
    \sigma^2\,\tr\big((\Sigmab_\mu+\lambda_n\I)^{-1}\big)\cdot
    \frac{1-\alpha_n}{d-n}\ +\
\frac{\w^{*\top}(\Sigmab_\mu+\lambda_n\I)^{-1}\w^*}
{\tr((\Sigmab_\mu+\lambda_n\I)^{-1})}\cdot (d-n),
&\text{for }n<d,\\
\sigma^2\, \tr(\Sigmab_\mu^{-1}),& \text{for }n=d,\\
\sigma^2\,\tr(\Sigmab_\mu^{-1})\cdot\frac{1-\beta_n}{n-d},&\text{for
}n>d,
\end{cases}
  \end{align*}
with $\lambda_n\geq 0$ defined by
  $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$, \
  $\alpha_n=\det(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$
  and $\beta_n=\ee^{d-n}$.
\end{theorem}
\begin{definition}
  We will use $\Mc(\Sigmab_\mu, \w^*,\sigma^2,n)$ to denote the above expressions
  for $\MSE{\Xb^\dagger\ybb}$.
\end{definition}

\begin{figure}[t]
\centering
\subfigure[%
  Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
  numerical estimates even for non-isotropic
  features. Eigenvalue decay leads to a steeper
  descent curve in the under-determined regime ($n<d$).]{%
    \includegraphics[width=0.48\textwidth]{figs/descent-intro}
  }
\hfill
\subfigure[%
    The mean of the estimator $\X^\dagger\y$ exhibits
    shrinkage which closely matches the shrinkage of a
    ridge-regularized least squares optimum (theory lines), as characterized by
    Theorem \ref{t:unbiased}.]{%
      \includegraphics[width=0.48\textwidth]{figs/descent-shrinkage}
    }
\caption{Illustration of the main results for $d=100$ and
$\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
eigenvalues decaying exponentially and scaled so that
$\tr(\Sigmab^{-1})=d$. We use our surrogate
formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
\ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
lines), accompanied by the empirical estimates based on the standard
i.i.d.~design (error bars are three times the standard error of the
mean). We consider three different condition numbers $\kappa$ of
$\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
i.e., $\Sigmab=\I$. We use $\sigma^2=1$ and
$\w^*=\frac1{\sqrt{d}}\one$.}
\vspace{-5mm}
\label{f:intro}
\end{figure}

\noindent
Proof of Theorem \ref{t:mse} is given in Appendix \ref{a:mse-proof}.
For illustration, we plot the MSE expressions in Figure~\ref{f:intro}a,
comparing them with empirical estimates of the true MSE under the
i.i.d.~design for a multivariate Gaussian distribution
$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
$100$ and vary the number of samples $n$, observing a double descent
peak at $n=d$. We observe that our theory aligns well with
the empirical estimates, whereas
previously, no such theory was available except for special
cases such as $\Sigmab=\I$ (more details in Theorem \ref{t:asymptotic}
and Appendices \ref{sec:proof-of-t-asymptotic} and \ref{sec:asymp-conj-details}). The plots
show that varying the spectral decay of $\Sigmab$ has a significant effect on the
shape of the curve in the under-determined regime. We use the
horizontal line to denote the MSE of the null estimator
$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
decay rapidly, then the Moore-Penrose estimator suffers less error
than the null estimator for some values of $n<d$, and the curve
exhibits a local optimum in this regime.

One important aspect of Theorem~\ref{t:mse} comes from the relationship between $n$ and the parameter $\lambda_n$, which together satisfy $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$.
This expression is precisely the classical notion of \emph{effective
  dimension} for ridge regression regularized with
$\lambda_n$~\citep{ridge-leverage-scores}, and it arises here even though there is
no explicit ridge regularization in the problem being considered in
Theorem~\ref{t:mse}. 
The global solution to the ridge regression task (i.e., $\ell_2$-regularized
least squares) with parameter $\lambda$ is defined as:
\begin{align*}
\argmin_\w \Big\{\E_{\mu,y}\big[\big(\x^\top\w-y(\x)\big)^2\big]
    + \lambda\|\w\|^2\Big\}\ =\ (\Sigmab_\mu +
  \lambda\I)^{-1}\v_{\mu,y},\quad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
\end{align*}
When Assumption \ref{a:linear} holds, then
$\v_{\mu,y}=\Sigmab_\mu\w^*$, however ridge-regularized least squares
is well-defined for much more general response models.
Our second result makes a direct connection between the (expectation
of the) unregularized minimum norm solution on the sample
and the global ridge-regularized solution.
While the under-determined regime (i.e., $n<d$) is of primary interest to us,
for completeness we state this result for arbitrary values of $n$ and $d$.
Note that, just like the definition of regularized least squares, this
theorem applies more generally than Theorem~\ref{t:mse}, in that it
does \emph{not} require the responses to follow any linear model as in
Assumption~\ref{a:linear} (proof in Appendix~\ref{s:unbiased-proof}). 
\begin{theorem}[Implicit regularization of Moore-Penrose estimator]
\label{t:unbiased}
For $\mu$ satisfying\footnote{The proof of Theorem \ref{t:unbiased}
  can be easily extended to probability measures $\mu$ that do not
  satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
  assumption here to simplify the presentation.} Assumption~\ref{a:general-position} and
  $y(\cdot)$ s.t.~$\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined, if
  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and $\yb_i=y(\xbb_i)$, then
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big] =
    \begin{cases}
       (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} &\text{for }n<d,\\
        \Sigmab_\mu^{-1}\v_{\mu,y} &\text{for }n \ge d,
    \end{cases}
  \end{align*}
  where, as in Theorem \ref{t:mse}, $\lambda_n$ is such that the effective dimension
  $\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$ equals $n$.
\end{theorem}

@@

We illustrate this result in Figure
\ref{f:intro}b, plotting the norm of the expectation of the
Moore-Penrose estimator. As for the MSE, our surrogate theory aligns
well with the empirical estimates for i.i.d.~Gaussian designs, showing
that the shrinkage of the unregularized estimator in the
under-determined regime matches the implicit
ridge-regularization characterized by Theorem \ref{t:unbiased}. While the shrinkage
is a linear function of the sample size $n$ for isotropic features
(i.e., $\Sigmab=\I$), it
exhibits a non-linear behavior for other spectral decays.
Such \emph{implicit regularization} has been studied
previously~\citep[see, e.g.,][]{MO11-implementing, %PM11, GM14_ICML,
Mah12}; it has
been observed empirically for RandNLA sampling
algorithms~\citep{MMY15}; and it has also received attention more
generally within the context of neural networks~\citep{Ney17_TR}. While our implicit regularization result
is limited to the Moore-Penrose estimator, this new connection (and
others, described below) between the minimum norm solution of an unregularized
under-determined system and a ridge-regularized least squares solution
offers a simple interpretation for the implicit regularization
observed in modern machine learning architectures.

Our exact non-asymptotic expressions in Theorem~\ref{t:mse} and
our exact implicit regularization results in Theorem~\ref{t:unbiased}
are derived for the surrogate design, but Figure \ref{f:intro}
suggests that they accurately describe the MSE (up to lower order
terms) also under the standard i.i.d.~design $\X\sim\mu^n$,
particularly when $\mu$ is a multivariate Gaussian.
As a third result, we can verify this in the cases where there exist
known expressions for the MSE under the i.i.d.~design (standard
Gaussian for the under-determined setting, and
arbitrary Gaussian for the over-determined one).
We give the proof in Appendix~\ref{sec:proof-of-t-asymptotic}.
\begin{theorem}[Asymptotic consistency of surrogate design]
\label{t:asymptotic}
Let $\rho=n/d\neq 1$, $\X\sim \mu^n$
and $y_i=y(\x_i)$ satisfy Assumption~\ref{a:linear}. If $d\geq
c_\rho=\frac3{|1-\rho|}$ and
\begin{align*}
  \mu &= \begin{cases}
    \Nc(\zero,\I)&\text{when }n<d-1,\\
    \Nc(\zero,\Sigmab),\ \Sigmab\succ\zero &\text{when }n>d+1,
  \end{cases}
\end{align*}
then the absolute difference between surrogate
expressions and the true MSE is bounded as follows:
% &\text{if}\qquad  (1)\quad \rho<1\ \text{ and }\ \mu=\Nc(\zero,\I)\qquad\text{or}\qquad
%     (2)\quad \rho > 1\ \text{ and }\ \mu=\Nc(\zero,\Sigmab),\
%     \Sigmab\succ\zero,\\
\begin{align*}
\Big|\MSE{\X^\dagger\y}-\Mc(\Sigmab, \w^*,\sigma^2,n)\Big|\ \leq\
\frac{c_\rho}{d} \cdot \Mc(\Sigmab,\w^*,\sigma^2,n).
\end{align*}
\end{theorem}
\begin{remark}
For $n$ equal to $d-1$, $d$ or $d+1$, the true MSE under Gaussian
random design can be infinite, whereas the
surrogate MSE is finite and has a closed form expression.
\end{remark}

\noindent
Empirical estimates given in Figure
\ref{f:intro} suggest that the consistency of surrogate expressions
holds much more generally than it is stated above. Based on a detailed
empirical analysis described in Appendix \ref{sec:asymp-conj-details},
we conjecture that an asymptotic consistency 
result of the form similar to the statement of Theorem~\ref{t:asymptotic}
holds true in the under-determined regime without the assumption that
$\Sigmab=\I$ (see Conjectures \ref{c:wishart} and
\ref{c:projection}). In this case, no formula is known for 
$\MSE{\X^\dagger\y}$, whereas the expressions for the surrogate
Gaussian design naturally extend.
